An Efficient Minibatch Acceptance Test for Metropolis-Hastings

### 2016-10-12 aistats2017.tex
Removed another space-consuming equation. OK, now it really is about 9.0 pages

Assume we begin with the current sample $\theta$ and a candidate sample
$\theta'$, and that $V \sim \mathcal{U}(0,1)$ is a uniform random variable. We
accept $\theta'$ if $g(\Delta(\theta,\theta')) > V$, and reject otherwise.
Since $g(s)$ is monotonically increasing, its inverse $g^{-1}(s)$ is
well-defined and unique. So an equivalent test is to accept $\theta'$ iff
\begin{equation}\label{eq:equivalent_test}
    \Delta(\theta,\theta') > X = g^{-1}(V)
\end{equation}
where $X$ is a random variable with the logistic distribution (its CDF is the
logistic function). To see this notice that $\frac{dV}{dX} = g'$, that $g'$ is
the density corresponding to a logistic CDF, and finally that $\frac{dV}{dX}$ is
the density of $X$. The density of $X$ is symmetric, so we can equivalently test
whether
\begin{equation}\label{eq:the_exact_test}
    \Delta(\theta,\theta') + X > 0
\end{equation}
for a logistic random variable $X$.
%{\color{blue} Daniel: I think the text
%before Equation~\ref{eq:the_exact_test} should be similar to what we had in our
%NIPS submission because I find it much clearer.}

@@

%% {\color{blue} Daniel: this subsection is improved from earlier. My main concern
%% now is that $X_{\rm norm}$ is described as being both approximately Gaussian
%% $\bar{\mathcal{N}}$ and Gaussian $\mathcal{N}$ here. We may want to clear that
%% up; in our NIPS submission we actually used a new variable $\varepsilon$ for
%% that and assigned it to be the $\mathcal{N}$ distribution. My current
%% suggestion is to have $\Delta^*$ be the $\bar{\mathcal{N}}$ approx. normal
%% distribution (but with mean $\Delta$, of course), and keep $X_{\rm norm}$ as an
%% exact Gaussian.}

@@

For computation of $C_\sigma$, we assume that its input $Y$ and another variable
$X$ lie in the intervals $[-V,V]$ and $[-2V,2V]$, respectively.  In continuous
form, the optimization problem is to find the $C_\sigma^*$ solving
\[
    \argmin_{C\sigma} \sup_{x \in [-2V,2V]}\left|\int_{-V}^{V}\Phi_{\sigma}(x-y) C_{\sigma}(y)dy - S(x)\right|.
\]
To get $C_\sigma^*$ in a computable form, we discretize $X$ and $Y$ into $4N+1$
and $2N+1$ values respectively. If $i \in \{-2N, \ldots, 2N\}=\mathcal{I}$ and
$j \in \{-N, \ldots, N\}$, then we can write $X_i = i(V/N)$ and $Y_j = j(V/N)$,
and the objective can be re-expressed as:
\[
    C_\sigma^* = \argmin_{C_\sigma} \max_{i \in \mathcal{I}}\left|\sum_{j = -N}^{N}\Phi_{\sigma}(X_i-Y_j) C_{\sigma}(Y_j) - S(X_i)\right|.
\]
We now define a matrix $M$ and vectors $u$ and $v$ such that $M_{ij} =
\Phi_{\sigma}(X_i-Y_j)$, $u_j = C_{\sigma}(Y_j)$, and $v_i = S(X_i)$, where the
indices $i$ and $j$ are appropriately translated to be non-negative indices for
$M, u,$ and $v$. Thus, the problem is now to minimize $\|Mu-v\|_{\infty}$ with
the constraint that $u > 0$ since it represents a density. We approximate this
with a least squares solution:
\begin{equation}\label{eq:optimization_l2}
    u^* = \argmin_u\; \|Mu-v\|_2^2 + \lambda \|u\|_2^2,
\end{equation}
where we add the standard regularization factor $\lambda$. The solution is
well-known from the normal equations: $u^* = (M^TM + \lambda I)^{-1}M^Tv$, and
in practice, it yields an acceptable $L_{\infty}$ norm.

@@

In the most similar prior works,~\citet{cutting_mh_2014} uses asymptotic
arguments based on the CLT to argue that its approximate acceptance test error
tends to zero as batch size increases, but no quantitative bounds are given.
In~\citet{icml2014c1_bardenet14}, explicit bounds are given, but they depend on
bounding:
\begin{equation}\label{eq:bad_bound}
    C_{\theta, \theta'} = \max_{1\leq i\leq N}|\log p(x_i\mid\theta') - \log p(x_i\mid\theta)|.
\end{equation}
Such bounds can be derived efficiently for models such as logistic regression,
but it is unclear how to derive them for a complex model such as a neural
network. In general, since a new $\theta'$ value is obtained each iteration, one
would need to iterate through all the $p(x_i\mid \theta')$ terms, defeating the
purpose of minibatch MCMC.

Here, we use quantitative forms of the Central Limit Theorem which rely on
measurable statistics from a single minibatch. Thus a sampler using our approach
does not need to see data beyond the current minibatch, and can easily be
applied on large datasets and in online modes where the data is presented as a
stream.

>>>>

Assume we begin with the current sample $\theta$ and a candidate sample
$\theta'$, and that $V \sim \mathcal{U}(0,1)$ is a uniform random variable. We
accept $\theta'$ if $g(\Delta(\theta,\theta')) > V$, and reject otherwise.
Since $g(s)$ is monotonically increasing, its inverse $g^{-1}(s)$ is
well-defined and unique. So an equivalent test is to accept $\theta'$ iff
\begin{equation}\label{eq:equivalent_test}
    \Delta(\theta,\theta') > X = g^{-1}(V)
\end{equation}
where $X$ is a random variable with the logistic distribution (its CDF is the
logistic function). To see this notice that $\frac{dV}{dX} = g'$, that $g'$ is
the density corresponding to a logistic CDF, and finally that $\frac{dV}{dX}$ is
the density of $X$. The density of $X$ is symmetric, so we can equivalently test
whether
\begin{equation}\label{eq:the_exact_test}
    \Delta(\theta,\theta') + X > 0
\end{equation}
for a logistic random variable $X$.

@@

@@

For computation of $C_\sigma$, we assume that its input $Y$ and another variable
$X$ lie in the intervals $[-V,V]$ and $[-2V,2V]$, respectively.
% Daniel: removing this because we were explaining this in quite some detail and
% it was taking a lot of space.
%\[
%    \argmin_{C\sigma} \sup_{x \in [-2V,2V]}\left|\int_{-V}^{V}\Phi_{\sigma}(x-y) C_{\sigma}(y)dy - S(x)\right|.
%\]
We discretize the convolution by discretizing $X$ and $Y$ into $4N+1$ and $2N+1$
values respectively. If $i \in \{-2N, \ldots, 2N\}=\mathcal{I}$ and $j \in \{-N,
\ldots, N\}=\mathcal{J}$, then we can write $X_i = i(V/N)$ and $Y_j = j(V/N)$, and the
objective can be written as:
\[
C_\sigma^* = \argmin_{C_\sigma} \max_{i \in \mathcal{I}}\left|\sum_{j\in\mathcal{J}} \Phi_{\sigma}(X_i-Y_j) C_{\sigma}(Y_j) - S(X_i)\right|.
\]
We now define a matrix $M$ and vectors $u$ and $v$ such that $M_{ij} =
\Phi_{\sigma}(X_i-Y_j)$, $u_j = C_{\sigma}(Y_j)$, and $v_i = S(X_i)$, where the
indices $i$ and $j$ are appropriately translated to be non-negative indices for
$M, u,$ and $v$. Thus, the problem is now to minimize $\|Mu-v\|_{\infty}$ with
the constraint that $u > 0$ since it represents a density. We approximate this
with a least squares solution:
\begin{equation}\label{eq:optimization_l2}
    u^* = \argmin_u\; \|Mu-v\|_2^2 + \lambda \|u\|_2^2,
\end{equation}
with regularization $\lambda$. The solution is well-known from the normal
equations ($u^* = (M^TM + \lambda I)^{-1}M^Tv$) and in practice yields an
acceptable $L_{\infty}$ norm.

@@

In the most similar prior works,~\citet{cutting_mh_2014} uses asymptotic
arguments based on the CLT to argue that its approximate acceptance test error
tends to zero as batch size increases, but no quantitative bounds are given.
In~\citet{icml2014c1_bardenet14}, explicit bounds are given, but they depend on
bounding:
\begin{equation}\label{eq:bad_bound}
    C_{\theta, \theta'} = \max_{1\leq i\leq N}|\log p(x_i\mid\theta') - \log p(x_i\mid\theta)|.
\end{equation}
Such bounds can be derived efficiently for models such as logistic regression,
but it is unclear how to derive them for a complex model such as a neural
network. In general, since a new $\theta'$ value is obtained each iteration, one
would need to iterate through all the $p(x_i\mid \theta')$ terms, defeating the
purpose of minibatch MCMC. In contrast, we use quantitative forms of the Central
Limit Theorem which rely on measurable statistics from a single minibatch. Thus
a sampler using our approach does not need to see data beyond the current
minibatch.
